Gaussian Processes Autoregressive Models for Forecasting the Disturbance Storm Time Index

### 2016-06-05 swsc.tex
Language changes proposed by Simon

   \author{M. Chandorkar
          \inst{1}
          \and
          S. Wing \inst{1}
          \and
          E. Camporeale\inst{1}
          }

@@

@@

Geomagnetic indices come in various forms, they may take continuous or discrete values, indices may be calculated at varying time frequencies. Their values are often calculated by averaging or combining a number of readings taken by instruments around the Earth. Each geomagnetic index is a proxy for a particular kind of phenomena. Some popular indices are the $K_p$, $Dst$ and the $AE$ index.

\begin{enumerate}
    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices, for a given value, for each of the past days. The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.

    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval. It is a continuous index which is calculated every hour.

    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents. 
\end{enumerate}

%Talk about Burton and friends
We focus on prediction of the hourly $Dst$ index which is a straightforward indicator of geomagnetic storms. More specifically, we focus on the \emph{one step ahead} (OSA) prediction of $Dst$ because it is the first piece towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $Dst$ OSA prediction problem has been the subject of several modeling efforts in the literature. One the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$.

@@

%Talk about neural networks
Yet another family of forecasting methods is based on \emph{Artificial Neural Networks} (ANN) that have been a popular choice for building predictive models. Researchers have employed both the standard \emph{feed forward} and the more specialized \emph{recurrent} architectures. \citet{Lund} proposed an \emph{Elman} recurrent network architecture called Lund $Dst$, which used the solar wind velocity, \emph{interplanetary magnetic field} (IMF) and historical $Dst$ data as inputs. \citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} [see \citet{boyle1997empirical}]. The same architecture is adapted for prediction of $Dst$ in \citet{SWE:SWE286}, which is popularly known as the Rice $Dst$ model. \citet{pallocchia:hal-00318011} proposed a \emph{neural network} model called EDDA to predict $Dst$ using only the IMF data.

@@

%Vision
In $Dst$ prediction a seemingly trivial yet highly informative prediction method is represented by the so called \emph{Persistence} model this is simply using the previous value of $Dst$ as the prediction for the next time step ($\hat{D}st(t) = Dst(t-1)$). Due to high correlation between $Dst$ values one hour apart ($Dst(t)$, $Dst(t-1)$), despite its apparent lack of sophistication the \emph{Persistence} model gives excellent predictive performance on error metrics. In essence it is a trivial hypothesis and hence it should always be used as a base line to compare the performance of any proposed $Dst$ algorithm. Moreover, given its zero computational cost, a model should at least outperform the \emph{Persistence} model in order to be a viable candidate for advancing the science of space weather prediction.

@@

Here $\matbf{f}$ is an $N\times 1$ vector consisting of the values $f(\mathbf{x}_i), i = q \cdots N$. In equation \ref{eq:normal} $\mathbf{f}|\mathbf{x}_1, \cdots, \mathbf{x}_N$ denotes the conditional distribution of $\mathbf{f}$ with respect to the input data (i.e. $\mathbf{X}$) and $\mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)$ represents a multivariate Gaussian distribution with mean vector $\mathbf{\mu}$ and covariance matrix $\mathbf{\Lambda}$. The probability density function of this distribution $p( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N)$ is therefore given by equation \ref{eq:sto}.

@@

Auto-regressive models can be improved by including exogenous quantities that can help in improving predictions. It is well known that the solar wind and interplanetary magnetic field are important factors that influence the Earth's geomagnetic state (\citet{JGRA:JGRA11775}, \citet{Russell1974}).

>>>>

   \author{M. Chandorkar
          \inst{1}
          \and
          E. Camporeale\inst{1}
          \and 
          S. Wing \inst{1}
          }

@@

   %\institute{Johns Hopkins University, Applied Physics Laboratory, \\ 11100 Johns Hopkins Road, Laurel, Maryland 20723}

@@

Geomagnetic indices come in various forms, they may take continuous or discrete values, indices may be calculated at varying time resolutions. Their values are often calculated by averaging or combining a number of readings taken by instruments around the Earth. Each geomagnetic index is a proxy for a particular kind of phenomena. Some popular indices are the $K_p$, $Dst$ and the $AE$ index.

\begin{enumerate}
    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices \cite{Bartels}. The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.

    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \cite{AEIndex}. It is a continuous index which is calculated every hour.

    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents and the strength of geomagnetic storms \cite{DesslerAndParker}. 
\end{enumerate}

%Talk about Burton and friends
For the present study, we focus on prediction of the hourly $Dst$ index which is a straightforward indicator of geomagnetic storms. More specifically, we focus on the \emph{one step ahead} (OSA), in this case one hour ahead prediction of $Dst$ because it is the first piece towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $Dst$ OSA prediction problem has been the subject of several modeling efforts in the literature. One the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where Q(t) relates to the particle injection from the plasma sheet into the inner magnetosphere.

@@

%Talk about neural networks
Yet another family of forecasting methods is based on \emph{Artificial Neural Networks} (ANN) that have been a popular choice for building predictive models. Researchers have employed both the standard \emph{feed forward} and the more specialized \emph{recurrent} architectures. \citet{Lund} proposed an \emph{Elman} recurrent network architecture called Lund $Dst$, which used the solar wind velocity, \emph{interplanetary magnetic field} (IMF) and historical $Dst$ data as inputs. \citet{JGRA:JGRA17461} used the recurrent neural networks to predict $Kp$. \citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} [see \citet{boyle1997empirical}]. The same architecture is adapted for prediction of $Dst$ in \citet{SWE:SWE286}, which is popularly known as the Rice $Dst$ model. \citet{pallocchia:hal-00318011} proposed a \emph{neural network} model called EDDA to predict $Dst$ using only the IMF data.

@@

%Vision
In $Dst$ prediction, a seemingly trivial yet highly informative prediction method is represented by the so called \emph{Persistence} model, which uses the previous value of $Dst$ as the prediction for the next time step ($\hat{D}st(t) = Dst(t-1)$). Due to high correlation between $Dst$ values one hour apart ($Dst(t)$, $Dst(t-1)$), despite its apparent lack of sophistication the \emph{Persistence} model gives excellent predictive performance on error metrics. In essence it is a trivial hypothesis and hence it should always be used as a base line to compare the performance of any proposed $Dst$ algorithm. Moreover, given its zero computational cost, a model should at least outperform the \emph{Persistence} model in order to be a viable candidate for advancing the science of space weather prediction.

@@

Here $\mathbf{f}$ is an $N\times 1$ vector consisting of the values $f(\mathbf{x}_i), i = q \cdots N$. In equation \ref{eq:normal} $\mathbf{f}|\mathbf{x}_1, \cdots, \mathbf{x}_N$ denotes the conditional distribution of $\mathbf{f}$ with respect to the input data (i.e. $\mathbf{X}$) and $\mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)$ represents a multivariate Gaussian distribution with mean vector $\mathbf{\mu}$ and covariance matrix $\mathbf{\Lambda}$. The probability density function of this distribution $p( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N)$ is therefore given by equation \ref{eq:sto}.

@@

Auto-regressive models can be improved by including exogenous quantities that can help in improving predictions. It is well known that the solar wind and interplanetary magnetic field are important factors that influence the Earth's geomagnetic state (\citet{JGRA:JGRA11775}, \citet{Russell1974}, \citet{JGRA:JGRA17574}).
